---
title: "HMM for Shark fisheries"
author: "Matias Braccini, and Marie-Pierre Etienne"
date: "April 16, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
list_used_packages <- c("depmixS4",  "tidyverse", "ggplot2", "ggmap")
## Be careful to laod dplyr after ggmap or use dplyr::summarize instead of summaryze
ind <- which(!list_used_packages %in% installed.packages())
list_packages_to_install <- list_used_packages[ind]
if (length(list_packages_to_install) > 0 ) install.packages(list_packages_to_install) 
invisible(lapply(list_used_packages , require, character.only = TRUE))
```


## First try with depmixS4 package

### Simulation of a HMM 

#### Simulation parameters

We set the parameters to some given values. The transistion matrix between states is defined by

```{r}
trans_mat <- matrix( c( .6, .3, 0, 0, .7, .6, .4, 0, .4), ncol = 3, nrow = 3)
```


We simulate some observe speed depending on the states

- State 1 corresponds to fishing = waiting. A toy example would be to have a log normal distribution with mean .1 and sd .15
- State 2 corresponds to retrieving the gear. mean 1 and sd .25
- State 3 corresponds to cruising. A mean 6 sd .4
```{r}
mu <- c(.1, 1, 6)
sigma <- c(.15, .25, .4)
```


We set the length of the sequence to 
```{r}
n <- 7 * 48
```

#### Simulation of the Hidden states

```{r}
hidden_states <- rep(NA, n)
hidden_states[1] <- 3

for( i in 2:n) {
  hidden_states[i] <- sample(1:3, size = 1, prob = trans_mat[hidden_states[i - 1],])
}
```


#### Simulation of the observed speed


```{r}
speed <- exp(rnorm(n, mean = log(mu[hidden_states]) - sigma[hidden_states]^2/2, sd = sigma[hidden_states]))
whole_data <- tibble(speed = speed, hidden = hidden_states, time = 1:n)

ggplot(whole_data, aes(x=time, y=speed, col= as.factor(hidden_states))) + geom_point() 

ggplot(whole_data, aes(x = speed, fill = as.factor(hidden_states)))+ geom_histogram() 

```


### Estimation  with no constraints

```{r}
mod <- depmix(speed  ~ 1, data = whole_data, nstates = 3, family = gaussian(log),
  transition =   ~ 1)
fm <- fit(mod)

summary(fm)
```


As the patterns are very clear the model has no issues  to estimate the transition

We can get the most credible sequence 

```{r}
posterior(fm)$state
```

and compare it to the 'true' states.

```{r}
table( posterior(fm)$state, whole_data$hidden )
```


## Test one nice trip with no constraint 

### Finding a good trip 

```{r}

vms_data <- read_csv('data/VMS.csv')
vms_data %>% dplyr::select( -AVERAGE_SPEED, -AVERAGE_COG, -Date.sent, -Year.sent, -Month.sent, -STATE_DATE, -Long.prev, -Lat.prev, -delta.pos, -VESSEL_TYPE) %>% mutate(Same.Trip_num = ifelse(Same.Trip == 'YES', 0, 1), trip_number = cumsum(Same.Trip_num) + 1)  %>% dplyr::select(-Same.Trip, -Same.Trip_num) -> small_vms_data


small_vms_data %>% group_by(trip_number) %>% summarise(n_pos_by_trip = n())

```


It seems to be always YES on the same trip variable .

I decide that a trip should not have 2 positions separetd by more than 10 hours.

```{r}
vms_data  %>%
  mutate(NAME= as.factor(NAME)) %>% 
  dplyr::select(-AVERAGE_SPEED, -AVERAGE_COG, -Date.sent, -delta.day.trip, Name.prev, -Month.sent, -STATE_DATE, -Long.prev, -Lat.prev, -delta.pos, -VESSEL_TYPE) %>% 
  dplyr::arrange(NAME, SENT_DATE_UTC)  %>% 
  group_by(NAME) %>%
  mutate(diff = SENT_DATE_UTC - lag(SENT_DATE_UTC, default = first(SENT_DATE_UTC))) %>% 
  ungroup() %>% 
  mutate(same_trip_bis = ifelse(diff < 3600*10, 0, 1), trip_number = cumsum(same_trip_bis) + 1, trip_id = factor(paste(NAME, trip_number, sep = '_' )))  -> sub_vms_data

```



We can focus on more recent data


```{r}
sub_vms_data  %>% filter(Year.sent  > 2015, diff < 3600) %>% group_by(trip_id) %>% summarise(n = n())
```


Let's try `Brooker_955'


```{r}
sub_vms_data  %>% filter(trip_id == 'Brooker_955') -> trip1_data
load(file = 'Australia.rda')
australia_map %>% ggmap( extent = "normal") +  geom_point(aes(x = LONGITUDE, y = LATITUDE), data = trip1_data, alpha = .5)
```


It seems to stay at the port.




Let's try `Corado_1880'


```{r}
sub_vms_data  %>% filter(trip_id == 'Corado_1880') -> trip1_data
trip1_data %>% summary()
load(file = 'Australia2.rda')
australia_map2 %>% ggmap( extent = "panel") +  geom_line(aes(x = LONGITUDE, y = LATITUDE), data = trip1_data, alpha = .5)
```


It seems better




